---
   title: "National Parks"
   author: "Joel Soroos"
   date: "September 29, 2019"
   output: pdf_document
---


### 1. Source data
```{r source, warning = TRUE, results = FALSE, message = FALSE}

   library (tidyverse)
   library (janitor) 

   visits_raw <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-09-17/national_parks.csv") %>%
      clean_names ()
```


### 2.  Transform visits
```{r transform, message = F, results = F}

   visits <- visits_raw %>%
      rename (park_name = unit_name) %>%
      filter (
         unit_type == "National Park",
         park_name != "Total"
         ) %>%
      select(year, park_name, state, visitors) %>%
      mutate (year = as.numeric (year)) %>%
      arrange (-year, -visitors)
```
```
### 3.  Calculate model
```{r}

   library (broom)

   combine <- loess(		    
          log10(cpu_actual) ~ year,		   
          data = cpu_max,		       
          na.action = na.omit
          ) %>%		
       augment (combine) %>%
       mutate (cpu_model = 10^.fitted)
```


### 4. Visualize data
```{r plot}

   library (scales)
   library (ggdark)

   ggplot () +
      #Lines
         #Moore's law prediction
             geom_line(
                data = combine,
                aes (x=year, y = cpu_moore),
                size = 0.4, linetype = 3
                ) +
         #Actual CPU model
            geom_line(
                data = combine,
                aes (x=year, y = cpu_model),
                size = 0.7, linetype = 2
                ) + 
      #Ribbons
         #Underpredict
            geom_ribbon (
               data = filter (combine, cpu_moore < cpu_model),
               aes (x=year, ymin = cpu_moore, ymax = cpu_model, fill = "Model fit > Moore's Law"),
               show.legend = T
               ) +
         #Overpredict
            geom_ribbon (
               data = filter (combine, cpu_moore > cpu_model),
               aes (x=year, ymin = cpu_moore, ymax = cpu_model, fill = "Model fit < Moore's Law"),
               show.legend = T
               ) +
      #Points
         geom_point(
            data = cpu_max,
            aes (x= year, y = cpu_actual, fill = "CPU actual"),
            size = 2, alpha = 0.6, show.legend = T
             ) +
      #Annotation
         #Moore's Law
             annotate(
                  geom="text", x=1991.5,y=30000000,
                  label="Moore's\nLaw", 
                  color="white", size=3, hjust=0.5,vjust=0.5, fontface="bold"
                  ) +
               geom_curve(
                  aes(x = 1991.5, xend = 1991.5, y = 13000000, yend = 3000000),
                  arrow = arrow(length = unit(0.1, "inch")),
                  size = 0.25, color = "white", curvature = 0  
                  ) +   
         #Model
                annotate(
                  geom="text", x=1990,y=80000,
                  label="Model", 
                  color="white", size=3, hjust=0.5,vjust=0.5, fontface="bold"
                  ) +
               geom_curve(
                  aes(x = 1990, xend = 1990, y = 99000, yend = 800000),
                  arrow = arrow(length = unit(0.1, "inch")),
                  size = 0.25, color = "white", curvature = 0  
                  ) +   
      #scales
         scale_y_continuous(
            trans = log10_trans(),
            breaks = c(2250, 10000, 100000, 1000000, 10000000, 100000000, 1000000000, 10000000000, 100000000000),
            labels = comma
            ) +
         scale_x_continuous(breaks = seq (1970, 2020, 10)) +
         scale_fill_manual(values = c("white","red", "green")) +
      #themes
         dark_theme_classic () +
         theme(
            plot.title = element_text(hjust = 0, vjust = 0, size = 17, face = "bold", margin = margin (0,0,7,0)),
            plot.title.position = "plot",
            plot.subtitle = element_text(hjust = 0, vjust = 0, size = 8, face = "bold", margin = margin (0,0,30,0)),
            plot.caption = element_text (hjust = 0, size = 8, face = "plain", margin = margin (15,0,0,0), color="#6D7C83"),
            plot.caption.position = "plot",
            panel.grid.major = element_line(colour="white", size=0.03),
            axis.title.y = element_text (size = 9, hjust = 0),
            axis.title.x = element_blank(),
            legend.position = c(0.87,0.33),
            legend.title = element_blank (),
            legend.text = element_text(size=8),
            legend.key.size = unit(0.2, "cm")
            ) +
         guides(fill = guide_legend(override.aes = list(shape = NA))) +
      labs(
         title = "Exponential growth in computing power",
         subtitle = "Gordon Moore predicted in 1971 that computing power would double approximately every 2 years (Moore's Law).\nOver the next 38 years he has largely been correct - slightly below through 2000, slightly above from 2001-2012 and slightly trailing more recently.",
         y = "# of transistors on a CPU microchip",
         caption = "Visualization: Joel Soroos @soroosj  |  Data: Wikipedia via R4DS Tidy Tuesday"
         ) +
      ggsave("moore.png", width = 22, height = 13, units = "cm")
```